Convergence rate of MAP estimates \\ for the exponential family

### 2021-01-12 main.tex
Updates

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    &\expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*)}\\
    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
    =& \bregman (\natp || \natp^*) \\
    =& \bregmanconj ( \meanp^* || \meanp) \\
    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
\begin{align}
	\label{eq:bregmanMLE}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \; , \\
	\label{eq:bregmanMAP}
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) \; .
\end{align}

@@

In the limit, the MAP estimates reach the Cramer-Rao lower bound. This can be seen by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* || \mu) 
    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
    + O(\norm{\mu - \mu^*}^3)\\
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where we introduced the Mahalanobis distance induced by the matrix 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1} 
    =: \mSigma^{-1}  \; .
\end{align}
To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
\begin{align}
	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
\end{align}
where $\mSigma$ is the covariance of $T(X)$. This is equal to $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$. It is worth $\nicefrac{d}{n}$ when $\mM = \mSigma^{-1}$. \RLP{Maybe that is Cramer-Rao, but check again.} And for the MAP we  get

\begin{align}
	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n}{(n+n_0)^2} \Tr(\mSigma^{-1}\mSigma) +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
Remark how the variance does not even appear, only the dimension divided by $n$ really matters. I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?

@@

In other words, we need for all $\lr\in(0,\inv{n})$, and all $\mu$ in the interior of the marginal polytope,

@@

	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 

@@

	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}

@@

\section{Variance bound for centered gaussians}
Sections on proximal point and mirror descent gave us the variance assumption 
\begin{align}
	\Cov(T(x_{n}) , \natp_{n+1}) = \Cov_X \bigg(T(X), \nabla \conj (\lr_n T(X) + (1-\lr_n)\mu_n)\bigg	) 
	\leq \lr_n M \; .
\end{align}
on the sufficient statistic and $\conj$. Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$ with natural parameter $\lambda > 0$ the precision and sufficient statistic $T(X)= - \half x^2$. The mean parameter is $- \half \sigma^2$ and  the conjugate map is $\nabla \conj (- \half \sigma^2) = \inv{\sigma^2} = \lambda$. 
\begin{align}
	\logpart (\lambda) = -\half \log(\lambda)  + \half \log \pi \\
	\conj(-\half \sigma^2) = -\half \log(\sigma^2) + \half (\log\frac{\pi}{2} - 1) \\
	\KL(\hat \sigma ; \sigma_*) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \sigma^2}
\end{align}

@@

\bibsection
\bibliographystyle{apalike}
\bibliography{references.bib}

>>>>

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    &\expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*)}\\
    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
    =& \bregman (\natp || \natp^*) \\
    =& \bregmanconj ( \meanp^* || \meanp) \\
    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
\begin{align}
	\label{eq:bregmanMLE}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) \; , \\
	\label{eq:bregmanMAP}
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right ) \; .
\end{align}

@@

In the limit, the MAP estimates reach the Cramer-Rao lower bound. This can be seen by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* || \mu) 
    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
    + O(\norm{\mu - \mu^*}^3)\\
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where we introduced the Mahalanobis distance induced by the matrix 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1} 
    =: \mSigma^{-1}  \; .
\end{align}
To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
\begin{align}
	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
\end{align}
where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
\begin{align}
	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
\RLP{Maybe that is Cramer-Rao, but check again.} And for the MAP we  get 
\begin{align}
	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
	&= \frac{n}{(n+n_0)^2} \Tr(\mSigma^{-1}\mSigma) +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
\end{align}
Remark how the variance does not even appear, only the dimension divided by $n$ really matters. I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?

@@

In other words, we need for all $\lr\in(0,\inv{n})$, and $ \mu$ in the interior of the marginal polytope,

@@

	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 

@@

	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}

@@

\section{Variance bound for centered gaussians}
Sections on proximal point and mirror descent gave us the variance assumption 
\begin{align}
	\Cov(T(x_{n}) , \natp_{n+1}) = \Cov_X \bigg(T(X), \nabla \conj (\lr_n T(X) + (1-\lr_n)\mu_n)\bigg	) 
	\leq \lr_n M \; .
\end{align}
on the sufficient statistic and $\conj$. Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$ with natural parameter $\lambda > 0$ the precision and sufficient statistic $T(X)= - \half x^2$. The mean parameter is $- \half \sigma^2$ and  the conjugate map is $\nabla \conj (- \half \sigma^2) = \inv{\sigma^2} = \lambda$. 
\begin{align}
	X \sim \cN(0,\sigma^2) \ ; \quad T(X)= - \half x^2 \ ; \quad  \mu = - \half \sigma^2  \\
	\natp = \lambda = \inv{\sigma^2} =  \nabla \conj (- \half \sigma^2) >0 \\
	\logpart (\lambda) = -\half \log(\lambda)  + \cst \\% \half \log \pi \\
	\conj(-\half \sigma^2) = -\half \log(\sigma^2) + \cst \\%\half (\log\frac{\pi}{2} - 1) \\
	\KL(\hat \sigma^2 ; \sigma_*^2) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \hat \sigma^2}
\end{align}

@@

\bibliographystyle{apalike}
\bibliography{references.bib}
